A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2016-06-29 rigid_bodies.tex
Changes in Section II - The title of Sec II.B was changed - Reference to Schay 1995 was added in the section - Derivatives dU/dr and dU/dq are now presented before the equation of motion - The equations of motion now have force and torque already substituted

One of the strategies employed to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations. Iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieved this suppression by adding constraints to the equations of motion. This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles. However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges a suitable approach. It is expected to be computationally more efficient because the constraints are treated implicitly. In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble. Being symplectic implies phase-space volume preservation and time reversibility, which are mandatory properties for numerical integrators to used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this sense, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions. This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide range of systems containing rigid molecules, including liquid water, \cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} etc. It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy often used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.

@@

This paper is organized as follows: in Sec.~\ref{sec:mechanics} we present a mathematical reformulation of the equations of motion followed by the Hamiltonian formalism. In Sec.~\ref{sec:moleculardynamics} we present the numerical integrators for both the NVE and NVT ensembles, whose performance we discuss in terms of the conservation of the total energy. Additionally, for the NVE case we present some results related to the partition of kinetic energy and for the NVT case, we also analyze the statistical distributions of the sampled energy. Following this, in Sec.~\ref{sec:pressure} we show the equation for the pressure developed in this work. Along this paper, we employ the word quaternion as a synonym for a vector in $R^4$ and avoid other interpretations. Only ordinary matrix/vector operations are present in the formulation. Some mathematical details are given in the Appendix.

@@

In this section we first review certain aspects of the kinematics or rigid bodies we consider relevant to this study. In Sec~\ref{sec:eulerparameters}, we introduce one particular factorization of the rotation matrix from which we obtain a mathematical relation between the rate of change of the four-dimensional quaternion and the angular velocity vector, while automatically satisfying the restriction imposed on the quaternion.

@@

Consider a rigid body composed of $n_p$ individual particles, each particle $j$ with mass $M_j$ and located at a Cartesian coordinate $\vt R_j$. The center of mass of this body is located at
\[
\vt r = \frac{1}{m}\sum_{j=1}^{n_p} M_j {\vt R}_j,
\]
where $m = \sum_{j=1}^{n_p} M_j$ is the total mass of the body. The position of each particle $j$ in a space-fixed reference frame whose origin coincides with the body's center of mass is ${\vt \delta}_j = \tr{ [\begin{array}{ccc} {\delta_j}_x & {\delta_j}_y & {\delta_j}_z \end{array}]} = {\vt R}_j - {\vt r}$. In such frame, the symmetric inertia tensor for rotations of the body about the origin is\cite{Goldstein2002}
\[
{\mt I}^\ast = \sum_{j=1}^N M_j \left[ \begin{array}{ccc}
{\delta_j}_y^2 + {\delta_j}_z^2 & -{\delta_j}_x {\delta_j}_y & -{\delta_j}_x {\delta_j}_z \\
-{\delta_j}_y {\delta_j}_x & {\delta_j}_x^2 + {\delta_j}_z^2 & -{\delta_j}_y {\delta_j}_z \\
-{\delta_j}_z {\delta_j}_x & -{\delta_j}_z {\delta_j}_y & {\delta_j}_x^2 + {\delta_j}_y^2 \\
\end{array} \right],
\]

For every non-collinear rigid body, there exists a particular frame of reference in which its inertia tensor is diagonal. The corresponding axes are known as the principal inertia axes and the formulation of the equations of motion in this frame of reference is notably simplified. Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\citep{Goldstein2002}
\[
{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \left[ \begin{array}{ccc}
I_1 &   0 &   0 \\
  0 & I_2 &   0 \\
  0 &   0 & I_3 \\
\end{array} \right],
\]
where $I_1$, $I_2$, and $I_3$ (the principal moments of inertia) are the eigenvalues of ${\mt I}^\ast$ and the columns of $\tr{\mt A}$ (the principal axes of rotation) are the corresponding eigenvectors. Algorithms for calculating $\mt A$ from $\mt I^\ast$ can be found, for example, in Ref.~\onlinecite{Kopp2008}. Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$. Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents a $k \times k$ identity matrix. If we differentiate both sides of this relation with respect to time, we conclude that $\mt A \tr{\dot {\mt A}} = - \dot{\mt A} \tr{\mt A}$, which means that $\mt A \tr{\dot{\mt A}}$ results in a skew-symmetric matrix. This product is related to $\vt \omega$, the angular velocity in the body-fixed frame, by\cite{Haug1989}
\begin{equation}
\label{eq:relation_A_omega}
\mt S(\vt \omega) = \mt A \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A},
\end{equation}
where $\mt S(\cdot)$ is the skew-symmetric operator defined as
\begin{equation}
\label{eq:operator_S}
\mt S(\vt x) = \left[ \begin{array}{ccc}
 0   & -x_3 &  x_2 \\
 x_3 &  0   & -x_1 \\
-x_2 &  x_1 &  0
\end{array}\right].
\end{equation}
The operator  $\mt S(\cdot)$ can be used to represent the cross-product of two vectors $\vt x$ and $\vt y$ as a matrix-vector product, since $\vt x \times \vt y = \mt S(\vt x)\vt y$.

%\subsection{Euler Parameters and Unit Quaternions}
\subsection{The Rate of change of Euler Parameters}
\label{sec:eulerparameters}

Euler's rotation theorem establishes that an arbitrary rotation can be described using three independent parameters. One can use, for instance, the so-called Euler angles, but mathematical singularities arise when the equations of motion are described in terms of those angles. Instead, we consider an alternative representation, in which the orientation of a rigid body is given by the Euler parameters $q_0$, $q_1$, $q_2$, and $q_3$, which form a quaternion $\vt q = \tr {[\begin{array}{cccc} q_0 & q_1 & q_2 & q_3 \end{array}]}$. The mathematical relationship between the Euler angles and the quaternion places the restriction over $\boldsymbol{q}$:\cite{Goldstein2002}
\begin{equation}
\label{eq:norm_eq_1}
\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_0^2 + q_1^2 + q_2^2 + q_3^2 = 1.
\end{equation}
Therefore, $\vt q$ is restricted to the surface of a unit hypersphere in $R^4$. Appendix~\ref{sec:quat_from_A} describes how one can compute the unit quaternion related to a given rotation matrix. Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\begin{widetext}
\begin{equation}
\label{eq:A_from_q}
\mt A = \left[ \begin{array}{ccc}
q_0^2 + q_1^2 - q_2^2 - q_3^2 & 2(q_1 q_2 + q_0 q_3) & 2(q_1 q_3 - q_0 q_2) \\
2(q_1 q_2 - q_0 q_3) & q_0^2 - q_1^2 + q_2^2 - q_3^2 & 2(q_2 q_3 + q_0 q_1) \\
2(q_1 q_3 + q_0 q_2) & 2(q_2 q_3 - q_0 q_1) & q_0^2 - q_1^2 - q_2^2 + q_3^2  
\end{array} \right].
\end{equation}
\end{widetext}

The rate of change of the Euler parameters is denoted by $\dot{\vt q}$. Differentiating Eq.~\ref{eq:norm_eq_1} with respect to time yields
\begin{equation}
\label{eq:diff_qTq}
\tr{\vt q}\dot{\vt q} = 0.
\end{equation}
This equation implies that, for Eq.~\ref{eq:norm_eq_1} to be preserved, $\dot{\vt q}$ must lie in the hyperplane orthogonal to $\vt q$.

Let us now introduce two new matrices that also depend on the Euler parameters, which are
\begin{subequations}
\label{eq:def_B_and_C}
\begin{align}
\mt B &= \left[
\begin{array}{rrrr}
-q_1 & -q_2 & -q_3 \\
 q_0 & -q_3 &  q_2 \\
 q_3 &  q_0 & -q_1 \\
-q_2 &  q_1 &  q_0
\end{array}
\right] \text{ and} \label{eq:def_B} \\
\mt C &= \left[
\begin{array}{rrrr}
-q_1 & -q_2 & -q_3 \\
 q_0 &  q_3 & -q_2 \\
-q_3 &  q_0 &  q_1 \\
 q_2 & -q_1 &  q_0
\end{array}
\right].
\end{align}
\end{subequations}
Some important properties of these matrices (or transposes thereof) have been reported in the literature\cite{Haug1989, Shuster1993, Dichmann1999, Ravishankar2004, Nielsen2012} and are revisited in Appendix \ref{sec:auxiliary_math} with our notation. The most remarkable feature is that
\begin{equation}
\label{eq:factorization_of_A}
{\mt A} = \tr{\mt B}{\mt C}.
\end{equation}
This is a simple and convenient factorization, since the entries of both $\mt B$ and $\mt C$ depend linearly on the Euler parameters. As demonstrated in Appendix~\ref{sec:auxiliary_math}, this factorization provides a relation between the angular velocity vector $\vt \omega$ and the rates of change of the Euler parameters. Such relation is
\begin{equation}
\label{eq:relation_qdot_omega}
\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
\end{equation}

@@

\subsection{Hamiltonian Dynamics}
\label{sec:hamiltonian}
In the previous subsection we obtained the relation between $\dot{\boldsymbol{q}}$ and $\boldsymbol{\omega}$ (Eq.~\ref{eq:relation_qdot_omega}), which is used in this part to develop the Hamiltonian formalism of the equations of motion. As usual, the dynamics of a rigid body can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively. The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$. Several authors have obtained Hamiltonian equations of motion for the rotation of rigid bodies using Euler parameters as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations.

We start by defining the Lagrangian $\mathcal{L} = K - U$ as a function of $\vt r$, $\dot{\vt r}$, $\vt q$, and $\dot{\vt q}$, where $K$ and $U$ are the kinetic and potential energies of the body, respectively. Since $K = 1/2 m \tr{\dot{\vt r}} \dot{\vt r} + 1/2 \tr{\vt \omega} \mt I \vt \omega$,\cite{Goldstein2002} we can resort to Eq.~\ref{eq:relation_qdot_omega} to write
\[
\mathcal{L} = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + 2 \tr{\dot{\vt q}} \mt B \mt I \tr{\mt B} \dot{\vt q} - U(\vt r, \vt q).
\]

To employ the Hamiltonian approach, besides the linear momentum $\vt p = \partial \mathcal{L}/\partial \dot{\vt r} = m \dot{\vt r}$, conjugated to $\vt r$, we introduce the quaternion momentum $\vt \pi = \partial \mathcal{L}/\partial \dot{\vt q}$, conjugated to $\vt q$.\citep{Goldstein2002} By performing the differentiation (see Appendix \ref{sec:Diff_Rules}), we get
\begin{equation}
\label{eq:conj_momentum}
\vt \pi = 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega.
\end{equation}
Hence, the conjugated momentum $\vt \pi$ is also orthogonal to $\vt q$. Using the relation above, we can write down the Hamiltonian $\mathcal{H} = K + U$ as a function of $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$, which is
\begin{equation}
\label{eq:H_with_B}
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + \frac{1}{8} \tr{\vt \pi} {\mt B} {\mt I}^{-1} \tr{\mt B} \vt \pi + U(\vt r, \vt q).
\end{equation}

The equations of motion are obtained from the Hamiltonian by $\dot{\vt r} = \partial \mathcal{H} / \partial \vt p$, $\dot{\vt p} = -\partial \mathcal{H} / \partial \vt r$, $\dot{\vt q} = \partial \mathcal{H} / \partial \vt \pi$, and $\dot{\vt \pi} = -\partial \mathcal{H} / \partial \vt q$.\cite{Goldstein2002} Eq.~\ref{eq:H_with_B} is convenient for differentiation with respect to $\vt \pi$, since $\mt B$ depends on $\vt q$ only. However, if we define
\[
\mt \Omega = \left[
\begin{array}{rrrr}
-\pi_1 & -\pi_2 & -\pi_3 \\
 \pi_0 & -\pi_3 &  \pi_2 \\
 \pi_3 &  \pi_0 & -\pi_1 \\
-\pi_2 &  \pi_1 &  \pi_0
\end{array}
\right],
\]
it follows that $\tr{\mt B}{\vt \pi} = -\tr{\mt \Omega}{\vt q}$ and, therefore,
\begin{equation}
\label{eq:H_with_Omega}
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + \frac{1}{8} \tr{\vt q} {\mt \Omega} {\mt I}^{-1} \tr{\mt \Omega} \vt q + U(\vt r, \vt q).
\end{equation}
Now this is convenient for differentiation with respect to $\vt q$.  Hence, after carrying out the differentiations as described in Appendix \ref{sec:Diff_Rules} and substituting $\vt \omega = (1/2) {\mt I}^{-1} \tr{\mt B} \vt \pi = (1/2) {\mt I}^{-1} \tr{\mt \Omega} \vt q$, we obtain the following system of differential equations to describe the dynamics of a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = -\diff{U}{\vt r} \\
&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \\
&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega - \diff{U}{\vt q}
\end{align}
\end{subequations}

For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference (see Appendix \ref{sec:Diff_PotEng}), and thus $\vt F$ is the resultant force on the body. To obtain the derivative $\partial U/\partial \vt q$, one must take into account that the entries of $\vt q$ are not independent, due to the constraint imposed by Eq.~\ref{eq:norm_eq_1}. As demonstrated in Appendix \ref{sec:Diff_PotEng}, we have
\[
\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed frame.

Anticipating a feature that will be convenient in the forthcoming sections, we now consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$ and according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
\begin{equation}
\label{eq:omega_entry}
\omega_k = \frac{\tr{\vt \pi} {\mt B}_k \vt q}{2 I_k},
\end{equation}
where $\mt B_k$ is the corresponding permutation matrix among those presented in Appendix \ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body becomes
\[
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
\]

>>>>

One of the strategies employed to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations. Iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieved this suppression by adding constraints to the equations of motion. This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles. However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges a suitable approach. It is expected to be computationally more efficient because the constraints are treated implicitly. In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble. Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions. This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide range of systems containing rigid molecules, including liquid water, \cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} etc. It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy often used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.

@@

This paper is organized as follows: in Sec.~\ref{sec:mechanics} we present a mathematical reformulation of the equations of motion of rigid bodies using a Hamiltonian formalism. In Sec.~\ref{sec:moleculardynamics} we devise numerical integrators for both the NVE and NVT ensembles, whose correctness we discuss in terms of conserved quantities. Additionally, for the NVE case we present some results related to kinetic energy partition and, for the NVT case, we also analyze the statistical distributions of the sampled energy. After that, in Sec.~\ref{sec:pressure} we present the equation developed for the pressure.

Note that, throughout this paper, we employ the word ``quaternion'' as a synonym for a four-dimensional vector and avoid other interpretations. Only ordinary matrix/vector operations are present in the formulation. Some mathematical details are given in the Appendix.

@@

In this section, we first review certain aspects of the kinematics of rigid bodies which are relevant to the present study. Unity quaternions are employed to represent the orientations of rigid bodies. Then, we introduce the use of a particular factorization of the rotation matrix, from which a mathematical relation arises between the rate of change of a four-dimensional quaternion and a three-dimensional angular velocity vector. An important feature of this relation is that it preserves the unit-norm constraint of the quaternion upon infinitesimal motion.

@@

Consider a rigid body composed of $n_p$ individual particles, with each particle $j$ having mass $M_j$ and being located at a Cartesian coordinate $\vt R_j$. The center of mass of this body is located at
\[
\vt r = \frac{1}{m}\sum_{j=1}^{n_p} M_j {\vt R}_j,
\]
where $m = \sum_{j=1}^{n_p} M_j$ is the total mass of the body. The position of each particle $j$ in a space-fixed frame of reference, whose origin coincides with the body's center of mass, is ${\vt \delta}_j = \tr{ [\begin{array}{ccc} {\delta_j}_x & {\delta_j}_y & {\delta_j}_z \end{array}]} = {\vt R}_j - {\vt r}$. In such frame, the symmetric inertia tensor for rotations of the body about the origin is\cite{Goldstein2002}
\[
{\mt I}^\ast = \sum_{j=1}^N M_j \left[ \begin{array}{ccc}
{\delta_j}_y^2 + {\delta_j}_z^2 & -{\delta_j}_x {\delta_j}_y & -{\delta_j}_x {\delta_j}_z \\
-{\delta_j}_y {\delta_j}_x & {\delta_j}_x^2 + {\delta_j}_z^2 & -{\delta_j}_y {\delta_j}_z \\
-{\delta_j}_z {\delta_j}_x & -{\delta_j}_z {\delta_j}_y & {\delta_j}_x^2 + {\delta_j}_y^2 \\
\end{array} \right].
\]

For every non-collinear rigid body, there exists a particular frame of reference in which its inertia tensor is diagonal. The corresponding axes are known as the principal axes and the formulation of the equations of motion in this frame of reference is notably simplified. Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\citep{Goldstein2002}
\[
{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \left[ \begin{array}{ccc}
I_1 &   0 &   0 \\
  0 & I_2 &   0 \\
  0 &   0 & I_3 \\
\end{array} \right],
\]
where $I_1$, $I_2$, and $I_3$ (the principal moments of inertia) are the eigenvalues of ${\mt I}^\ast$ and the columns of $\tr{\mt A}$ (the principal axes of rotation) are their corresponding eigenvectors. Algorithms for calculating $\mt A$ from $\mt I^\ast$ can be found, for example, in Ref.~\onlinecite{Kopp2008}. Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$. Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents an identity matrix in $\mathbb{R}^{k \times k}$ . By differentiating both sides of this relation with respect to time, we conclude that $\mt A \tr{\dot {\mt A}} = - \dot{\mt A} \tr{\mt A}$, which means that $\mt A \tr{\dot{\mt A}}$ results in a skew-symmetric matrix. This product is related to $\vt \omega$, the angular velocity of the body in the body-fixed frame, by\cite{Haug1989}
\begin{equation}
\label{eq:relation_A_omega}
\mt S(\vt \omega) = \mt A \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A},
\end{equation}
where $\mt S(\cdot)$ is the skew-symmetric operator defined as
\begin{equation}
\label{eq:operator_S}
\mt S(\vt x) = \left[ \begin{array}{ccc}
 0   & -x_3 &  x_2 \\
 x_3 &  0   & -x_1 \\
-x_2 &  x_1 &  0
\end{array}\right].
\end{equation}

The operator  $\mt S(\cdot)$ is able to represent the cross-product of two vectors $\vt x$ and $\vt y$ as a matrix-vector product, since $\vt x \times \vt y = \mt S(\vt x)\vt y$. This property is explored in the Appendix.

%\subsection{Euler Parameters and Unit Quaternions}
\subsection{The Rate of Change of Unit Quaternions}
\label{sec:eulerparameters}

Euler's rotation theorem establishes that an arbitrary rotation can be described using three independent parameters. One can use, for instance, the so-called Euler angles, but mathematical singularities arise when the equations of motion are described in such terms. Instead, we consider an alternative representation, in which the orientation of a rigid body is given by the Euler parameters $q_0$, $q_1$, $q_2$, and $q_3$, which form a quaternion $\vt q = \tr {[\begin{array}{cccc} q_0 & q_1 & q_2 & q_3 \end{array}]}$. The mathematical relation between the Euler angles and the Euler parameters requires that\cite{Goldstein2002}
\begin{equation}
\label{eq:norm_eq_1}
\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_0^2 + q_1^2 + q_2^2 + q_3^2 = 1.
\end{equation}

Therefore, $\vt q$ is restricted to the surface of a unit hypersphere in $\mathbb{R}^4$, which is itself a three-dimensional domain. Appendix~\ref{sec:quat_from_A} describes how one can compute the unit quaternion related to a given rotation matrix. Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\begin{widetext}
\begin{equation}
\label{eq:A_from_q}
\mt A = \left[ \begin{array}{ccc}
q_0^2 + q_1^2 - q_2^2 - q_3^2 & 2(q_1 q_2 + q_0 q_3) & 2(q_1 q_3 - q_0 q_2) \\
2(q_1 q_2 - q_0 q_3) & q_0^2 - q_1^2 + q_2^2 - q_3^2 & 2(q_2 q_3 + q_0 q_1) \\
2(q_1 q_3 + q_0 q_2) & 2(q_2 q_3 - q_0 q_1) & q_0^2 - q_1^2 - q_2^2 + q_3^2  
\end{array} \right].
\end{equation}
\end{widetext}

The rate of change of a quaternion is denoted by $\dot{\vt q}$. Differentiating Eq.~\ref{eq:norm_eq_1} with respect to time yields
\begin{equation}
\label{eq:diff_qTq}
\tr{\vt q}\dot{\vt q} = 0,
\end{equation}
which implies that, for Eq.~\ref{eq:norm_eq_1} to be preserved upon an infinitesimal motion, $\dot{\vt q}$ must lie in the hyperplane orthogonal to $\vt q$.

Let us now introduce two new matrices that also depend on the Euler parameters, which are
\begin{subequations}
\label{eq:def_B_and_C}
\begin{align}
\mt B &= \left[
\begin{array}{rrrr}
-q_1 & -q_2 & -q_3 \\
 q_0 & -q_3 &  q_2 \\
 q_3 &  q_0 & -q_1 \\
-q_2 &  q_1 &  q_0
\end{array}
\right] \text{ and} \label{eq:def_B} \\
\mt C &= \left[
\begin{array}{rrrr}
-q_1 & -q_2 & -q_3 \\
 q_0 &  q_3 & -q_2 \\
-q_3 &  q_0 &  q_1 \\
 q_2 & -q_1 &  q_0
\end{array}
\right].
\end{align}
\end{subequations}

Some important properties of these matrices (or transposes thereof) have been reported in the literature\cite{Haug1989, Shuster1993, Dichmann1999, Ravishankar2004, Nielsen2012} and are revisited with our particular notation in Appendix~\ref{sec:auxiliary_math}. The most remarkable feature is that
\begin{equation}
\label{eq:factorization_of_A}
{\mt A} = \tr{\mt B}{\mt C}.
\end{equation}

The equation above is a simple and convenient factorization of matrix $\mt A$, since the entries of both $\mt B$ and $\mt C$ depend linearly on the Euler parameters. As demonstrated in Appendix~\ref{sec:auxiliary_math}, this factorization provides a relation between the angular velocity vector $\vt \omega$ and the rates of change of the Euler parameters. Such relation is
\begin{equation}
\label{eq:relation_qdot_omega}
\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
\end{equation}

@@

\subsection{Hamiltonian Dynamics}
\label{sec:hamiltonian}

The relation between $\dot{\boldsymbol{q}}$ and $\boldsymbol{\omega}$ (Eq.~\ref{eq:relation_qdot_omega}) obtained in the previous subsection is used here to develop a Hamiltonian formalism for the motion of rigid body. As usual, it can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively. The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$. Several authors have obtained Hamiltonian equations for the rotation of rigid bodies using Euler parameters as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations.

We start by defining the Lagrangian $\mathcal{L} = K - U$ as a function of $\vt r$, $\dot{\vt r}$, $\vt q$, and $\dot{\vt q}$, where $K$ and $U$ are the kinetic and potential energies of the body, respectively. Since $K = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + \frac{1}{2} \tr{\vt \omega} \mt I \vt \omega$,\cite{Goldstein2002} we can resort to Eq.~\ref{eq:relation_qdot_omega} to write
\[
\mathcal{L} = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + 2 \tr{\dot{\vt q}} \mt B \mt I \tr{\mt B} \dot{\vt q} - U(\vt r, \vt q).
\]

To employ the Hamiltonian approach, besides the linear momentum $\vt p = \partial \mathcal{L}/\partial \dot{\vt r} = m \dot{\vt r}$, conjugated to $\vt r$, we introduce the quaternion momentum $\vt \pi = \partial \mathcal{L}/\partial \dot{\vt q}$, conjugated to $\vt q$.\citep{Goldstein2002} By performing the differentiation (see Appendix \ref{sec:Diff_Rules}), we get
\begin{equation}
\label{eq:conj_momentum}
\vt \pi = 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega.
\end{equation}

Hence, the conjugated momentum $\vt \pi$ is also orthogonal to $\vt q$ (premultiplication of $\mt I \vt \omega$ by $\mt B$). Using the relation above, we can write down the Hamiltonian $\mathcal{H} = K + U$ as a function of $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$, which is
\begin{equation}
\label{eq:H_with_B}
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + \frac{1}{8} \tr{\vt \pi} {\mt B} {\mt I}^{-1} \tr{\mt B} \vt \pi + U(\vt r, \vt q).
\end{equation}

The equations of motion are obtained from the Hamiltonian by $\dot{\vt r} = \partial \mathcal{H} / \partial \vt p$, $\dot{\vt p} = -\partial \mathcal{H} / \partial \vt r$, $\dot{\vt q} = \partial \mathcal{H} / \partial \vt \pi$, and $\dot{\vt \pi} = -\partial \mathcal{H} / \partial \vt q$.\cite{Goldstein2002} Eq.~\ref{eq:H_with_B} is convenient for differentiation with respect to $\vt \pi$, since $\mt B$ depends on $\vt q$ only. However, if we define
\[
\mt \Omega = \left[
\begin{array}{rrrr}
-\pi_1 & -\pi_2 & -\pi_3 \\
 \pi_0 & -\pi_3 &  \pi_2 \\
 \pi_3 &  \pi_0 & -\pi_1 \\
-\pi_2 &  \pi_1 &  \pi_0
\end{array}
\right],
\]
it follows that $\tr{\mt B}{\vt \pi} = \tr{\mt \Omega}{\vt q}$ and, therefore, a convenient form of $\mathcal{H}$ for differentiation with respect to $\vt q$ is
\begin{equation}
\label{eq:H_with_Omega}
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + \frac{1}{8} \tr{\vt q} {\mt \Omega} {\mt I}^{-1} \tr{\mt \Omega} \vt q + U(\vt r, \vt q).
\end{equation}

For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference (see Appendix \ref{sec:Diff_PotEng}), and thus $\vt F$ is the resultant force on the body. To obtain the derivative $\partial U/\partial \vt q$, one must take into account that the entries of $\vt q$ are not independent, due to the constraint imposed by Eq.~\ref{eq:norm_eq_1}. The correct way of performing differentiation with respect to a constrained vector is described in Ref.~\onlinecite{Schay1995}. As demonstrated in Appendix \ref{sec:Diff_PotEng}, we have
\[
\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed reference frame.

Finally, after carrying out the needed differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\vt \omega = (1/2) {\mt I}^{-1} \tr{\mt B} \vt \pi = (1/2) {\mt I}^{-1} \tr{\mt \Omega} \vt q$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \mt F \\
&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \\
&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau
\end{align}
\end{subequations}

Anticipating a feature that will be convenient in the forthcoming sections, we now consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
\begin{equation}
\label{eq:omega_entry}
\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
\end{equation}
where $\mt B_k$ is the corresponding permutation matrix among those presented in Appendix \ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body becomes
\[
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
\]
